Occultation mapping of Io's surface in the near-infrared I: Inferring static maps

### 2020-12-28 paper/paper.tex
Reformatting the paper with the split into paper 1/2 in mind.

% Title
%\title{Inferring a time dependent map of Io's surface from occultations and phase curves.}
\title{Time-variable surface mapping of exoplanets: an application to Io}

@@

\begin{abstract}
Jupiter's moon Io is the most volcanically active body in the Solar System with hundreds of active volcanoes varying in intensity on different timescales.
Existing photometric observations in the near infrared taken during occultations by Jupiter and Galilean moons encode a wealth of information about its surface.
Io is thus an ideal testbed for models of time-variable exoplanet surfaces.
We built a generative model of Io's surface for each observation using Nonnegative Matrix Factorization.
The model is built on top of the code starry \href{https://rodluger.github.io/starry/}{\color{linkcolor}\faGithub} which enables fast analytic computation of occultation light curves, phase curves in both emitted and reflected light and exact gradients with respect to all input parameters.
We use the inferred map to study the time variability of prominent volcanoes over a timescale of decades.
While we mostly focus on Io, our models are directly applicable to mapping exoplanets.
We discuss this application towards the end of the paper.\href{https://github.com/fbartolic/volcano}{\color{linkcolor}\faGithub}

\end{abstract}

%
\section{Introduction}
%% ADD PARAGRAPH REVIEWING PREVIOUS LITERATURE ON MAPPING EXOPLANETS, MENTION JWST AND LUVOIR
%% MAYBE EVEN IN THE ABSTRACT
The surface of Jupiter's moon Io is littered with hundreds of volcanoes appearing as bright spots in the near infrared and varying in intensity on timescales of days to decades.
The volcanic activity is driven by tidal interactions with Jupiter and sustained by the Laplace resonance with Europa and Ganymede \citep{peale_melting_1979}.
Io is a perfect analogue of an extremely volcanically active exoplanet.
Such rocky exoplanets, often dubbed \emph{super-Ios} or \emph{lava worlds}, have gathered a lot of interest in recent years because the photometric and spectral signatures of volcanic activity on such worlds will likely be detectable in the near future \citep{kaltenegger_detecting_2010,henning_highly_2018,oza_sodium_2019}.
Studying volcanism on both Io and exoplanets is scientifically valuable because it provides a window into the properties of early Earth and it informs theories of planet formation.

Existing exoplanet detections with potential volcanism include CoRoT-7b \citep{barnes_corot-7b_2010}, the first discovered rocky exoplanet with likely significant tidal heating; 55 Cancri e, whose longitudinal offset in peak surface emission has been attributed to, among other things, lava flows on the surface \citep{demory_variability_2016,demory_map_2016,hammond_linking_2017}; several planets in the TRAPPIST-1 system in a Laplace-like resonance with likely volcanic activity \citep{kislyakova_magma_2017,dobos_tidal_2019}; and many others.
Not all exoplanets will have Io-like volcanism, some will have magma oceans due to their proximity to the star or volcanism caused by nuclear decay similar to early Earth volcanism.

Io has been observed extensively using both space and ground observatories.
High resolution images of Io's surface were taken by space missions such as Voyager \citep{smith_jupiter_1979}, Galileo \citep{belton_galileos_1996} and Juno \citep{mura_infrared_2020}.
It has also been resolved from the ground in the near infrared using disk resolved imaging
\citep{howell_infrared_1985,simonelli_disk-resolved_1986,spencer_discovery_1990} and adaptive optics observations \citep{marchis_adaptive_2000,marchis_keck_2005,de_kleer_spatial_2016}.
Most importantly for this work, Io has been sporadically observed for decades using high cadence near infrared photometry taken during occultations by Jupiter starting with \cite{spencer_discovery_1990}.
Occultations by Jupiter occur twice every orbit lasting ~1.7 days and mutual occultations with Europa, Ganymede or Callisto happen every ~6 years when Earth pasess through the orbital plane of Galilean satellites.
The majority of occultations by Jupiter are observed when Io is in Jupiter's shadow ("in eclipse") while mutual occultations are almost always observed in sunlight because mutual occultations while Io is in eclipse are extraordinarily rare.
Only the brightest volcanoes are visible over the reflected light background when Io is illuminated by the Sun\citep{veeder_ios_1994,de_kleer_spatial_2016}.

Both kinds of occultations have been used to study volcanic activity on the surface.
\cite{spencer_io_1994} observed several occultations of Io by Europa and detected a major brightening of the most powerful of Io's volcanoes, Loki, relative to previous Voyager observations.
\cite{rathbun_loki_2002, rathbun_loki_2006,rathbun_ground-based_2010} used observations from NASA's IRTF telescope to study the long term variability of different volcanic spots finding evidence of periodicity in Loki's eruptions and established the transient nature of the volcanoes (only Loki is persistently active).
By studying an occultation of Io by Europa, \cite{de_kleer_multi-phase_2017} mapped the Loki region to a precision of about 2 kilometers.
In addition to observations of occultations in near infrared, several groups have been observing mutual occultations in the optical \citep[][and references therein]{arlot_mutual_1974,saquet_phemu15_2018,morgado_astrometry_2016} with the purpose of inferring the optical albedo of Io and improving ephemeris precision for Galilean satellites.

None of the studies of occultations in the infrared attempted to infer a two-dimensional map of Io's surface.
The light curves are usually fit independently with the goal of inferring one-dimensional longitudinal variations in brightness on its surface.
Strong priors on the locations of individual volcanoes are assumed based on maps from space mission observations.
In this work, we build a fully probabilistic model in order to infer a time-dependent two-dimensional \emph{map} of Io's volcanic surface from archival IRTF observations in the near infra red.
The model is built on top of the recently developed code starry \citep[][Luger et al. 2020 in prep]{luger_starry_2019} which
enables fast analytic computation of occultation light curves and phase curves for objects whose surface features are represented in terms of spherical harmonics.
starry can compute phase curves and occultations in both emitted light (for modeling the isotropic volcanic emission from Io's surface) and reflected light (for mapping albedo variations).
Instead of fitting one map per occultation light curve, we reduce the dimensionality of the problem using matrix factorization, assuming a fixed map for a given light curve which is a linear combination of $K$ basis maps.
We use priors on the map features which ensure the maps are physical and prevent overfitting of the data.
The model is interpretable, each of the basis maps roughly corresponds to a single spot on the surface, and the coefficients multiplying the basis maps encode information on the time variability.

Our goals for this paper are to both improve on past studies of Io using occultation light curves and to construct a flexible model directly applicable to future efforts on inferring time-dependent maps of variable features on exoplanet surfaces such as volcanoes, magma oceans or clouds.
The fact that we have lots of prior information on what the surface of Io should look like enables us to test our model in a realistic setting, testing sensitivity to chosen priors, data quality, the orbital parameters and different kinds of surface features.

The paper is organized as follows.
In \S\ref{sec:data} we describe the IRTF light curves of occultations by Jupiter which we use to infer maps.
In \S\ref{sec:model} we discuss the generative model for the data for the case of a static map, and a dynamic map modeled using a matrix factorization approach and we write down the relevant likelihood functions.
In \S\ref{sec:inverse_problem} we focus on the question of how to do Bayesian inference given the forward model specifications defined in \S\ref{sec:model}.
We discuss and test different priors on map features, the information content of light curves and fitting the models using variational inference (VI) methods and Hamiltonian Monte Carlo (HMC).
We present results on simulated data.
\S\ref{sec:results} shows the results on real data.
In particular, the inferred dynamic map, the  different component maps and accompanying coefficients, the uncertainties of the maps, sensitivity to different choices of priors, comparison of inferred locations and intensities of spots to known values from the literature.
Finally, in \S\ref{sec:exoplanets} we discuss applications of the model to future exoplanet obervations and test the model predictions on simulated JWST observations of a volcanic Super-Earth.

@@

\section{Model}
\label{sec:model}
\subsection{Orbital parameters}
\label{ssec:orbital_parameters}
To model the occultations we need to know the projected separation between Jupiter and Io and the orientation of Io at any given time.
We use the \href{https://ssd.jpl.nasa.gov/horizons.cgi}{JPL Horizons database} and the latest planetary ephemeris DE430 \citep{folkner_planetary_2014} which are accurate to a few kilometers on Io's surface.
Specifically, we use the Right Ascension and Declination, the angular (equatorial) diameter (\textsf{Ang-diam}), the visibility flag indicating whether Io is occulted or not and whether it is in partial or total eclipse (\textsf{Ang-sep/v}), the longitude and latitude at the center of Io's disc as seen from Earth (\textsf{Ob-lon}, \textsf{Ob-lat}), the sub-solar point longitude and latitude which specify the direction pointing towards the Sun on Io's surface (\textsf{Sl-lon}, \textsf{Sl-lat}), the distance from Io to the Sun (\textsf{r}) and
the counterclockwise angle between the Celestial North Pole unit vector projected onto the plane of the sky and the Io's north pole (\textsf{NP.ang}).
All longitudes are positive to the west.
\textsf{HORIZONS} provides all ephemeris with a minimum cadence of 1 second so we interpolate all values such that we can evaluate them for arbitrary times.

@@

\subsection{Nonnegative Matrix Factorization}
\label{ssec:nmf}
In the case when we cannot assume a static map, the situation is considerably more complicated.
In principle, the surface map is different for each \emph{data point}.
Of course, fitting one map per data point is intractable so we might instead fit a single map per light curve, generated by Eq.~\ref{eq:linear_model_pix}.
Even then, we would need to fit on the order of 1k parameters!
This is both becuse the number of spherical coefficients scales as $(l+1)^2$ and because when fitting in pixel space we need many more pixels than spherical harmonics to ensure positivity.
Although fitting a model of such high dimensionality for a single IRTF light curve of high signal to noise ratio is tractable thanks to exact gradients provided by autodifferentiation in \textsf{starry}, we don't want to do that for a hundred or so separate light curves.
Such a model would be very difficult to fit and it would require strong regularization between successive maps in time.
It also wouldn't directly provide much physical insight into the volcanic activity, we would have to conduct a separate analysis on the inferred maps to constrain the time variability of the volcanoes.
Instead, we need an approach which reduces the dimensionality of the problem.
One way of accomplishing that is to expand the spherical harmonic coefficients (or pixels) in a Taylor series in time about a certain point.
This is the approach \cite{luger_tess_2019} took to model a map of Earth's albedo in reflected light from stray Earthshine in the aperture of the TESS space telescope.
This issue with this approach and similar series expansions such as the Fourier series is that with the very long baseline of IRTF data and the observations being scattered throughout the years sporadically, it is unclear which point should be the origin of the expansion.
More importantly, there likely doesn't exist a well defined global time-averaged map of Io so requiring too much smoothness between successive maps wouldn't work.
Another way of reducing the dimensionality of the problem would be a parametric approach where where we place a few spots on the sphere and then fit only for the locations and intensities of each spot.
This is assuming that we know a priori that the surface features are spot like and roughly how many we should expect.
While such a strong assumption might work for studying individual spots in the context of Io, it certainly isn't justified for exoplanets.
In Section~\ref{sec:inverse_problem} we do assume stronger priors on the locations of and shapes of surface features to better constrain the variability of individual volcanoes, but we still fit for pixels, at least in principle allowing the data to override our assumptions.

The approach approach opted for is to treat the problem of inferring a single map per light curve as a probabilistic matrix factorization problem.
The idea is to assume that a map for any given light curve can be expressed as a linear combination of $K$ "basis maps".
Consider $L$ lightcurves, the model for the $l$-th light curve is
\begin{equation}
    \mathbf{f}_l=\mathbf{A}_l\,\mathbf{P}^\dagger\,\mathbf{p}_l'
\end{equation}
We can stack the column vectors $\mathbf{p}_l$ into a matrix $\mathbf{Y}$ of shape $(N_p, L)$ where $N_p$ is the number of pixels for each light curve.
We then assume that the $\mathbf{Y}$ can be decomposed into a product of two matrices $\mathbf{B}$ and $\mathbf{Q}$ as
\begin{equation}
    \mathbb{Y}=\mathbf{B}\,\mathbf{Q}
    \label{eq:nmf}
\end{equation}
where $\mathbf{B}$ has shape $(N_p, K)$ and $\mathbf{Q}$ has shape $(K, L)$.
The model for all light curves can be written as
\begin{equation}
    \mathbb{f}=\mathbb{A}\,\mathrm{vec}(\mathbf{\mathbb{Y}})
    \label{eq:model_all_lcs}
\end{equation}
Where $\mathbf{f}$ is a tall column vector consisting of predictions for all light curves stacked together, $\mathbb{A}$ is a block diagonal matrix with matrices $\mathbf{A}_l\,\mathbf{P}^\dagger$ on the diagonal for $l=1\dots, L$ and the $\mathrm{vec}$ operator stacks the columns of $\mathbb{Y}$ into a tall vector.
The model is \emph{bilinear} which means that is linear in the matrices $\mathbf{B}$ and $\mathbf{Q}$ separately.
The interpretation of Equation~\ref{eq:nmf} is simple, each of the $K$ columns of $\mathbf{B}$ represents pixels of a basis map and the columns of $\mathbf{Q}$ determine how those maps add together to produce a final map for the $l$-th light curve.
For this reason we call $\mathbf{B}$ the \emph{basis matrix} and the matrix $\mathbf{Q}$ the \emph{encoding matrix}.
Ideally, we want the basis maps to be physically meaningful and the coefficients encode the time variability.
A nice feature of this model is that there is no requirement that successive maps are smoothly varying between different light curves.

The matrix factorization problem as written in Equation~\ref{eq:nmf} is highly degenerate and practically intractable a probabilistic framework with a small dataset.
Since each of the columns of $\mathbf{B}$ are pixels representing emitted light or albedo, we can substantially reduce the ambiguity in the decomposition by requiring that both matrices are strictly positive.
This matrix factorization problem is known as \emph{Nonnegative Matrix Factorization} (NMF) \citep{paatero_positive_1994,lee_algorithms_2001} and it has a rich history across many different fields.
It is commonly used for decomposing physical signals in the spectral or the time domain.
Even though NMF is more tractable than unconstrained matrix factorization, it is still an NP hard problem \citep{vavasis_complexity_2009} and it requires additional constraints or priors to be tractable.
Simultaneously transforming $\mathbf{B}\leftarrow \mathbf{B}\,\mathbf{S}^{-1}$ and
$\mathbf{Q}\leftarrow \,\mathbf{S}\mathbf{Q}$ with a nonsingular matrix $\mathbf{S}$ doesn't change the value of the objective function.
The objective function is also invariant to a simultaneous permutation of the columns of $\mathbf{B}$ and rows of $\mathbf{Q}$ and rescaling either of the two matrices by a scalar.
Nevertheless, getting around these degeneracies in practice is possible, especially when we have a lot of prior knowledge on the problem.
For a recent review article with a detalied analysis of these degeneracies and common algorithms, see \cite{fu_nonnegative_2019}.

Broadly, there are two approaches to fitting NMF models, the constrained optimization approach and the probabilistic approach.
In the optimization approach we optimize $\mathbb{Y}$ in a least-squares sense under a set of constraints on either one or both of the matrices \citep[see][for recent examples from the astronomical literature]{acosta-pulido_new_2017, melchior_scarlet_2018,ren_non-negative_2018,ren_using_2020} with the goal of obtaining a point estimate for the two matrices.
The probabilistic approach introduces priors and the inference is usually done with variational inference in order to obtain an approximate posterior distribution over the matrices.
In this work we opt for the probabilistic approach because we care about the uncertainties on the map features a lot.

One difference between most applications of NMF in the literature and NMF within the context of inferring surface maps from 1D light curves is that
in the majority of the work in the literature \citep[except][]{kawahara_global_2020} the matrix $\mathbb{Y}$ is assumed to be directly observed  up to a simple noise term.
In our case the problem is considerably more challenging because we observe the light curves
$\mathbb{A}\,\mathrm{vec}(\mathbf{\mathbb{Y}})$ instead of $\mathbb{Y}$ directly, a process in which some information is lost because, depending on the geometry of the occultations, certain combinations of spherical harmonic coefficients will be in the nullspace.
\cite{kawahara_global_2020} are using NMF to model phase variations in directly imaged exoplanets with the goal of simultaneously inferring a surface map and a spectral decomposition of the map into several components, a case in which the
matrix which is to be decomposed is also unobserved.
Although we don't attempt to solve the problem of spectrally decomposing the maps, our approach is directly applicable to that problem as well. (TODO: should I add more details here?)
One could also imagine inferring a multi spectral component and time dependent map by solving two NMF problems simultaneously.
We leave this application for future work.

\subsection{The likelihood}
\label{ssec:likelihood}
Given Eq.~\ref{eq:nmf} we can write down the likelihood for our data and in Section~\ref{sec:inverse_problem} we dicuss in detail the priors which make the problems of inferring static and dynamic maps tractable.
Assuming a Gaussian noise process for the data with a dense covariance matrix, the (log) likelihood is
\begin{equation}
    \ln\mathcal{L}=-\frac{1}{2}\left[\mathbb{f}_\mathrm{obs}-(\mathbb{f} + \mathbb{b})\right]^{\top} \boldsymbol{\Sigma}^{-1}\left[\mathbb{f}_\mathrm{obs}-(\mathbb{f} + \mathbb{b})\right]
\end{equation}
where $\mathbb{f}_\mathrm{obs}$ is the vector of stacked observed light curves, $\boldsymbol{\Sigma}$ is the data covariance matrix and $\mathbb{b}$ is a fixed flux offset per light curve accounting for stray flux not attributed to Io (mostly due to Jupiter).
We model the data covariance with a Gaussian Process using the fast Celerite method\citep{foreman-mackey_fast_2017} as implemented in the \textsf{exoplanet} package (TODO: cite) to capute correlations imposed by the seeing plus an additional white noise term which is different for each light curve.
The IRTF data isn't provided with any uncertainties so we estimate the uncertainties by filtering the data with the Savitzky Golay filter as implemented in \textsf{SciPy} (TODO: CITE) and estimating the variance of the residuals.
We assume identical errors for all data points in a given light curve but allow for a rescaling factor when fitting the model.
In the next section, we add priors to this likelihood and fit the model on simulated data using optimization and Variational Inference (VI).

@@

% bulk of the results in the paper
\subsection{The information content of a light curve}
\label{ssec:information_content}
The mapping problem is famously ill posed.
In other words, certain linear combinations of spherical harmonic coefficients will be in the nullspace of the linear mapping  $\mathbf{A}$ in Eq.~\ref{eq:linear_model}.
This means that even if we had noiseless observations, certain features would still be impossible to recover.
To recover surface features from observed flux integrated over the disc of the object we need
to have some mechanism which breaks the various degeneracies.
For phase curves in emitted light, we can only recover longitudinal variations in intensity.
Occultations are a lot better thanks to the limb of the occultor sweeping across the surface of the occulted sphere.
Even better are grazing occultations with an occultor of a small diameter sweeping across the disc at different latitudes and therefore breaking the latitudinal degeneracies. 
Observing phase curves and occultations in reflected light improves the model even more because of the nonuniform illumination profile of the incident radiation and the presence of a day/night terminator line \citep[][Luger et al. 2020 in prep]{luger_starry_2019}.
In some cases for reflected light observations (phase curves of an inclined planet for example) there is no nullspace at all at low degrees of the map.

@@

We take the ephemeris data  from \textsf{JPL HORIZONS} and assume that it is known exactly, we also assume all observations are of the emitted light component independent of whether Io is in sunlight or in eclipse.
This is because we are interested in constraining the volcanic emission rather than the albedo.
Given the fixed design matrices, we can compute the variance reduction of the posterior relative to the prior known as \emph{posterior shrinkage}, defined as
\begin{equation}
    1-\frac{\sigma^2_\mathrm{post}}{\sigma^2_\mathrm{prior}}
\end{equation}
where $\sigma^2_\mathrm{prior}$ is the prior and $\sigma^2_\mathrm{post}$ is the posterior variance for a given parameter.
To compute this quantity for each spherical harmonic coefficient in $\mathbf{y}$ we divide the diagonal elements of matrices $\boldsymbol{\Sigma}_{\hat{\mathbf{y}}}$ and $\boldsymbol{\Lambda}$, divide them, and average over all $m$-modes. 
We need to assume some variance for the Gaussian prior on the $Y_{l,m}$ coefficients so we choose a broad prior $\boldsymbol{\Lambda}=\mathrm{diag}(1^2,0.1^2,\dots,0.1^2)$.
Although the posterior shrinkage depends on this prior the main results do not change for a different choice of the prior.
The result is plotted in Fig.~\ref{fig:information_content}.
\begin{figure}[h!]
    \begin{centering}
    \includegraphics[width=0.5\linewidth]{figures/information_content.pdf}
    \oscaption{InformationContent}{
        The plot shows the information content of different kinds of observations of Io as a function of spherical harmonic degree (angular scale of surface features).
        On the y axis is posterior shrinkage (the variance ratio between the prior and the posterior distribution) for each spherical harmonic coefficient at a given degree $l$ on the x axis, averaged across all $m$ modes.
        Posterior shrinkage of 1 represent maximum information gain in updating from the prior to the posterior while a shrinkage of 0 represents no information gain. 
        We compute the posterior variance for different kinds of simulated observations of Io over
        the course of a single year: phase curves (blue), occultations by Jupiter (orange), combined phase curves and occultations by Jupiter (green) and occulltations of Io by other Galilean moons (red lines). 
        Solid lines represent observations at SNR=100 while dashed lines represent observations at SNR=10.
        \label{fig:information_content}
    }
    \end{centering}
\end{figure}
The plot shows posterior shrinkage for phase curve observations (blue lines), occultations by Jupiter (orange), the previous two combined (green) and mutual occultations by other Galilean moons (red). 
Solid lines correspond to SNR=100 while dashed lines correspond to SNR=10 where the signal is defined as the min-max difference of the predicted flux.

@@

% Bibliography
\bibliography{bib}
\end{document}

>>>>

% Title
\title{Occultation mapping of Io's surface in the near-infrared I: Inferring static maps}

@@

\begin{abstract}
Jupiter's moon Io is the most volcanically active body in the Solar System with hundreds of active volcanoes varying in intensity on different timescales.
Existing photometric observations in the near infrared taken during occultations by Jupiter and other Galilean moons encode a wealth of information about its surface.
These data are ideal for testing methods for inferring surface features of occulted bodies from time series observations.
We built a generative model of Io's surface using the code starry \href{https://rodluger.github.io/starry/}{\color{linkcolor}\faGithub} which enables fast analytic and differentiable computation of occultation light curves in emitted and reflected light.
    Our (fully Bayesian) model is able to recover known bright features on the surface of Io with only two light curves and without any assumptions on where the features are located or how many there are.
    This work is a first step towards constructing a model for time-variable surface features. 
    Although we do not present new findings about Io in this work, the methods we have developed are directly applicable to the problem of mapping stars and exoplanets.\href{https://github.com/fbartolic/volcano}{\color{linkcolor}\faGithub}
\end{abstract}

%
\section{Introduction}
The surface of Jupiter's moon Io is littered with hundreds of volcanoes which appear as bright spots in the near infrared and in intensity on timescales ranging from days to decades.
The volcanic activity is driven by tidal interactions with Jupiter and sustained by the Laplace resonance with Europa and Ganymede \citep{peale1979}.
Because of its extreme volcanic activity, Io is in many ways an analogue of a volcanically active exoplanet.
Volcanic exoplanets, often dubbed \emph{super-Ios} or \emph{lava worlds}, have gathered a lot of interest in recent years because both photometric and spectral signatures of volcanic activity on such worlds will likely be detectable in the near future with telescopes such as JWST and LUVOIR \citep{kaltenegger2010,henning2018,oza2019}.
Studying extreme volcanism on both Io and exoplanets is scientifically valuable because it provides a window into the properties of the early Earth and it informs theories of planet formation.

Existing exoplanet detections with potential volcanism include CoRoT-7b \citep{barnes2010}, the first rock exoplanet to have been discovered and one which likely heated by strong tidal forces; 55 Cancri e, whose inferred longitudinal offset in peak surface emission has been attributed to, among other things, lava flows on the surface \citep{demory2016,demory2016a,hammond2017}; several planets in the TRAPPIST-1 system in a Laplace-like resonance with likely volcanic activity \citep{kislyakova2017,dobos2019} and many others.
Not all exoplanets will have Io-like volcanism, some will have magma oceans due to their proximity to the star or volcanism caused by nuclear decay similar to early Earth volcanism.

Io has been observed extensively using both space and ground observatories.
High resolution images of Io's surface were taken by space missions such as Voyager \citep{smith1979}, Galileo \citep{belton1996} and Juno \citep{mura2020}.
It has also been resolved from the ground in the near infrared using disk resolved imaging
\citep{howell1985,simonelli1986,spencer1990} and adaptive optics observations \citep{marchis2000,marchis2005,dekleer2016}.
Most importantly for this work, Io has been sporadically observed for decades using high cadence near infrared photometry taken during occultations by Jupiter starting with \cite{spencer1990}.
Occultations by Jupiter occur twice every orbit lasting ~1.7 days and mutual occultations with Europa, Ganymede or Callisto happen every ~6 years when Earth pasess through the orbital plane of Galilean satellites.
The majority of occultations by Jupiter are observed when Io is in Jupiter's shadow ("in eclipse") while mutual occultations are almost always observed in sunlight because mutual occultations while Io is in eclipse are extraordinarily rare.
Only the brightest volcanoes are visible over the reflected light background when Io is illuminated by the Sun\citep{veeder1994,dekleer2016a}.

Both kinds of occultations have been used to study volcanic activity on the surface.
\cite{spencer1994} observed several occultations of Io by Europa and detected a major brightening of the most powerful of Io's volcanoes, Loki, relative to previous Voyager observations.
\cite{rathbun2002, rathbun2006,rathbun2010} used observations from NASA's IRTF telescope to study the long term variability of different volcanic spots finding evidence of periodicity in Loki's eruptions and established the transient nature of the volcanoes (only Loki is persistently active).
By studying an occultation of Io by Europa, \cite{dekleer2017} mapped the Loki region to a precision of about 2 kilometers.
In addition to observations of occultations in near infrared, several groups have been observing mutual occultations in the optical \citep[][and references therein]{arlot1974,saquet2018,morgado2016a} for decades with the purpose of inferring the optical albedo of Io and improving ephemeris precision for Galilean satellites.

None of the studies of occultations in the infrared attempted to infer a two-dimensional map of Io's surface.
The light curves are usually fit independently with the goal of inferring one-dimensional longitudinal variations in brightness on its surface.
Strong priors on the locations of individual volcanoes are assumed based on maps from space mission observations.
In this work, we build a fully probabilistic model in order to infer a two-dimensional \emph{map} of Io's volcanic surface from archival IRTF observations in the near infra red.
We assume that the map generating a given set of light curves is static (up to an overall amplitude) so we are restricted to modeling the surface during a timespan when the surface emission hasn't changed drastically.
In a subsequent paper we build on the model we have developed here and generalize to a case where the surface is time-variable.

The model is built on top of the recently developed code starry \citep[][Luger et al. 2020 in prep]{luger2019a} which enables fast analytic computation of occultation light curves and phase curves for objects whose surface features are represented in terms of spherical harmonics.
Starry can compute phase curves and occultations in both emitted light (for modeling the isotropic thermal emission from Io's surface) and reflected light (for mapping albedo variations).
It is many orders of magnitude faster and more accurate than pixel based codes and it computes exact gradients with respect to all parameters through automatic differentiation.
Starry also comes with extensive tools for visualizing spherical harmonic maps and simulating data.

Our goal for this paper is to develop a general model for mapping surfaces of occulted bodies with sparse high contrast features and to test it on a real world dataset. 
Because we have lots of prior information on what the surface of Io should look like, we can compare our inferences to known features on its surface and we can test the sensitivity of the inferred maps to different choices of priors and noise models.
The work should also be of interest to the Io research community as it presents a very different approach to modeling Io occultations without having to make strong assumptions on the locations of the hot spots.

The paper is organized as follows.
In \S\ref{sec:data} we describe the IRTF light curves of occultations by Jupiter which we use to infer maps.
In \S\ref{sec:model} we discuss the generative model for the data and we write down the relevant likelihood functions.
In \S\ref{sec:inverse_problem} we focus on the question of how to do Bayesian inference given the forward model specifications defined in \S\ref{sec:model}.
We discuss and test different priors on map features, the information content of light curves and fitting the modelsusing Hamiltonian Monte Carlo (HMC). 
We show results on simulated data.
\S\ref{sec:results} shows the results on real IRTF data for two pairs of light curves (closely spaced in time)
In particular, the inferred dynamic map, the  different component maps and accompanying coefficients, the uncertainties of the maps, sensitivity to different choices of priors, comparison of inferred locations and intensities of spots to known values from the literature.
Finally, in \S\ref{sec:exoplanets} we discuss applications of the model to future exoplanet obervations and test the model predictions on simulated JWST observations of a volcanic Super-Earth.

@@

\section{Model}
\label{sec:model}
\subsection{Orbital parameters}
\label{ssec:orbital_parameters}
To model the occultations we need to know the projected separation between Jupiter and Io and the orientation of Io at any given time.
We use the \href{https://ssd.jpl.nasa.gov/horizons.cgi}{JPL Horizons database} and the latest planetary ephemeris DE430 \citep{folkner2014} which are accurate to a few kilometers on Io's surface.
Specifically, we use the Right Ascension and Declination, the angular (equatorial) diameter (\textsf{Ang-diam}), the visibility flag indicating whether Io is occulted or not and whether it is in partial or total eclipse (\textsf{Ang-sep/v}), the longitude and latitude at the center of Io's disc as seen from Earth (\textsf{Ob-lon}, \textsf{Ob-lat}), the sub-solar point longitude and latitude which specify the direction pointing towards the Sun on Io's surface (\textsf{Sl-lon}, \textsf{Sl-lat}), the distance from Io to the Sun (\textsf{r}) and
the counterclockwise angle between the Celestial North Pole unit vector projected onto the plane of the sky and the Io's north pole (\textsf{NP.ang}).
All longitudes are positive to the west.
\textsf{HORIZONS} provides all ephemeris with a minimum cadence of 1 second so we interpolate all values such that we can evaluate them for arbitrary times.

@@

\subsection{The likelihood}
\label{ssec:likelihood}
Given Eq.~\ref{eq:nmf} we can write down the likelihood for our data and in Section~\ref{sec:inverse_problem} we dicuss in detail the priors which make the problems of inferring static and dynamic maps tractable.
Assuming a Gaussian noise process for the data with a dense covariance matrix, the (log) likelihood is
\begin{equation}
    \ln\mathcal{L}=-\frac{1}{2}\left[\mathbb{f}_\mathrm{obs}-(\mathbb{f} + \mathbb{b})\right]^{\top} \boldsymbol{\Sigma}^{-1}\left[\mathbb{f}_\mathrm{obs}-(\mathbb{f} + \mathbb{b})\right]
\end{equation}
where $\mathbb{f}_\mathrm{obs}$ is the vector of stacked observed light curves, $\boldsymbol{\Sigma}$ is the data covariance matrix and $\mathbb{b}$ is a fixed flux offset per light curve accounting for stray flux not attributed to Io (mostly due to Jupiter).
We model the data covariance with a Gaussian Process using the fast Celerite method\citep{foreman-mackey2017} as implemented in the \textsf{exoplanet} package (TODO: cite) to capute correlations imposed by the seeing plus an additional white noise term which is different for each light curve.
The IRTF data isn't provided with any uncertainties so we estimate the uncertainties by filtering the data with the Savitzky Golay filter as implemented in \textsf{SciPy} (TODO: CITE) and estimating the variance of the residuals.
We assume identical errors for all data points in a given light curve but allow for a rescaling factor when fitting the model.
In the next section, we add priors to this likelihood and fit the model on simulated data using optimization and Variational Inference (VI).

@@

% bulk of the results in the paper
\subsection{The information content of a light curve}
\label{ssec:information_content}
The mapping problem is famously ill posed.
In other words, certain linear combinations of spherical harmonic coefficients will be in the nullspace of the linear mapping  $\mathbf{A}$ in Eq.~\ref{eq:linear_model}.
This means that even if we had noiseless observations, certain features would still be impossible to recover.
To recover surface features from observed flux integrated over the disc of the object we need
to have some mechanism which breaks the various degeneracies.
For phase curves in emitted light, we can only recover longitudinal variations in intensity.
Occultations are a lot better thanks to the limb of the occultor sweeping across the surface of the occulted sphere.
Even better are grazing occultations with an occultor of a small diameter sweeping across the disc at different latitudes and therefore breaking the latitudinal degeneracies. 
Observing phase curves and occultations in reflected light improves the model even more because of the nonuniform illumination profile of the incident radiation and the presence of a day/night terminator line \citep[][Luger et al. 2020 in prep]{luger2019a}.
In some cases for reflected light observations (phase curves of an inclined planet for example) there is no nullspace at all at low degrees of the map.

@@

We take the ephemeris data  from \textsf{JPL HORIZONS} and assume that it is known exactly, we also assume all observations are of the emitted light component independent of whether Io is in sunlight or in eclipse.
This is because we are interested in constraining the volcanic emission rather than the albedo.
Given the fixed design matrices, we can compute the variance reduction of the posterior relative to the prior known as \emph{posterior shrinkage}, defined as
\begin{equation}
    1-\frac{\sigma^2_\mathrm{post}}{\sigma^2_\mathrm{prior}}
\end{equation}
where $\sigma^2_\mathrm{prior}$ is the prior and $\sigma^2_\mathrm{post}$ is the posterior variance for a given parameter.
To compute this quantity for each spherical harmonic coefficient in $\mathbf{y}$ we divide the diagonal elements of matrices $\boldsymbol{\Sigma}_{\hat{\mathbf{y}}}$ and $\boldsymbol{\Lambda}$, divide them, and average over all $m$-modes. 
We need to assume some variance for the Gaussian prior on the $Y_{l,m}$ coefficients so we choose a broad prior $\boldsymbol{\Lambda}=\mathrm{diag}(1^2,0.1^2,\dots,0.1^2)$.
Although the posterior shrinkage depends on this prior the main results do not change for a different choice of the prior.
The result is plotted in Fig.~\ref{fig:information_content}.
\begin{figure}[h!]
    \begin{centering}
    \includegraphics[width=0.5\linewidth]{figures/information_content.pdf}
    \oscaption{InformationContent}{
        The information content of different kinds of observations of Io as a function of spherical harmonic degree (angular scale of surface features).
        On the y axis is posterior shrinkage (the variance ratio between the prior and the posterior distribution) for each spherical harmonic coefficient at a given degree $l$ on the x axis, averaged across all $m$ modes.
        Posterior shrinkage of 1 represent maximum information gain in updating from the prior to the posterior while a shrinkage of 0 represents no information gain. 
        We compute the posterior variance for different kinds of simulated observations of Io over
        the course of a single year: phase curves (blue), occultations by Jupiter (orange), combined phase curves and occultations by Jupiter (green) and occulltations of Io by other Galilean moons (red lines). 
        Solid lines represent observations at SNR=100 while dashed lines represent observations at SNR=10.
        \label{fig:information_content}
    }
    \end{centering}
\end{figure}
The plot shows posterior shrinkage for phase curve observations (blue lines), occultations by Jupiter (orange), the previous two combined (green) and mutual occultations by other Galilean moons (red). 
Solid lines correspond to SNR=100 while dashed lines correspond to SNR=10 where the signal is defined as the min-max difference of the predicted flux.

@@

% Bibliography
\bibliography{bib}

\end{document}
